### 2014-04-04 HW_non-modal.tex
April 4

Normal mode analysis -- the calculation of eigenvalues and eigenvectors of a linearized dynamical system -- has been used to solve many problems over the years.
Despite its wide-ranging success, it has failed in important instances, particularly in predicting the onset of subcritical turbulence in hydrodynamic flows. 
The reason for this failure was explained in the early 1990's when Trefethen and others attributed the pitfalls of normal mode analysis to the non-normality of linear operators of
dynamical systems~\cite{trefethen1993,schmid2007}. A non-normal operator has 
eigenvectors that are not orthogonal to one another. One consequence of eigenvector nonorthogonality is that even when all eigenvectors decay exponentially under linear evolution, 
superpositions of eigenvectors can grow, albeit transiently.
In other words, certain fluctuations of the laminar state can access free energy from background gradients even though normal mode fluctuations cannot.
When combined with nonlinear effects, this allows for sustained subcritical turbulence.
Such behavior is obscured by traditional normal mode analysis, which only effectively describes the long time asymptotic behavior of fluctuations under  
action of the linear operator. Transient growth, which can dominate turbulent evolution, can be discovered only through non-modal calculations.

Non-modal analysis has been embraced by the hydrodynamics community over the past two decades in the attempt to explain and predict subcritical turbulence. But the plasma community
generally relies on normal mode analysis to inform turbulent predictions and observations, with a few exceptions~\cite{camargo1998,camporeale2010,schekochihin2012}. 
Furthermore, non-modal treatments have generally been explanatory rather than predictive and have centered around the transition to turbulence in subcritical systems rather 
than on properties of fully-developed turbulence.
This paper takes up the task of developing an approach to understand turbulent properties using only non-modal linear calculations with the goal of 
ultimately making quantitative predictions.  Our approach is to calculate an average growth rate of turbulent fluctuations due to linear processes, specifically due to transient growth.  We model the
turbulent steady state as a series of processes:  (1) the turbulence starts as a spatially random state, (2) linear transient growth deterministically amplifies the turbulent energy (or
decrease it in wavenumber ranges where linear damping dominates), (3) nonlinear transfer sets in at a specified timescale, terminating the
transient growth process and re-randomizing the turbulent state (at which point the cycle repeats).    Optimally, the timescale for the final step would be the
nonlinear decorrelation time of the turbulent system, but in order to enable predictive capability, we employ critical balance
arguments to use a characteristic \emph{linear} time. Since there is no obvious single linear time scale, we test several and compare the results to determine which works best.
The procedure ultimately produces a growth rate that can be used
to predict turbulent properties such as saturation levels and transport rates through mixing length arguments.

@@

\beqar
\label{n_eq}
\pdiff{n}{t} = - {\mathbf v_E} \cdot \grad n - \kappa \pdiff{\phi}{y} + \xi \gradpar^2 (n - \phi) - D \gradperp^4 n, \\
\label{phi_eq}
\pdiff{\gradperp^2 \phi}{t} = - {\mathbf v_E} \cdot \grad (\gradperp^2 \phi) + \xi \gradpar^2 (n - \phi) - D \gradperp^6 \phi
\eeqar
The normalizations are $\phi \to e \phi/T_e, n \to n/n_0, \grad \to \rho_s \grad , x,y,z \to x,y,z/\rho_s, t \to \omega_{ci} t $. Additionally,
$\mathbf{v_E} = \mathbf{b} \times \grad \phi$, $\xi = 4 \pi \omega_{ce}/\nu_{ei}$, 
and $\kappa = \rho_s/L_n$, where $L_n$ is the density scale length. The terms $D \gradperp^4 n$ and $D \gradperp^6 \phi$ are artificial hyper-diffusion and hyper-viscosity used to dissipate
energy on small scales. We set $D \to 10^{-5}$ except when otherwise indicated.
We associate the direction along $\mathbf{B}$ as the parallel or $z$ direction. The direction perpendicular $\mathbf{B}$ that
points opposite the density gradient is defined as the $x$ direction, and the $y$ direction is taken perpendicular to both of these. 
In the analysis and numerical simulations, which we perform with the BOUT++ code~\cite{dudson2009}, we use periodic boundary conditions in
the $y$ and $z$ directions with zero-value boundaries in the $x$ direction. For the 2D version of the model, we replace $-\xi \gradpar^2$ by 
$\alpha = k_z^2 \xi = 4 \pi k_z^2 \rho_s^2 \omega_{ce} /\nu_{ei}$, which is commonly known as the adiabaticity parameter~\cite{camargo1995,camargo1998}.

@@

\beqar
\label{fourier_eqn}
\frac{\partial}{\partial t} \left( \begin{array}{cc} n_k \\ \phi_k \end{array} \right) = \mathbf{A}_k \left( \begin{array}{cc} n_k \\ \phi_k \end{array} \right) + \mathbf{N}_k \left( \begin{array}{cc} n_k \\ \phi_k \end{array} \right), \\ \nonumber \\
\mathbf{A}_k = \left( \begin{array}{cc} -\xi k_z^2 - D k_\perp^4 & -i \kappa k_y + \xi k_z^2 \\  \xi \frac{k_z^2}{k_\perp^2} & - \xi \frac{k_z^2}{k_\perp^2} - D k_\perp^4\end{array} \right)
\eeqar
with $\mathbf{A}_k$ being the linear coupling matrix and $\mathbf{N}_k$ the nonlinear matrix, which we don't write explicitly. 
Now in assessing the degree of normality of the HW model, we are not interested in the normality of $\mathbf{A}_k$ itself. 
Rather, we are interested in the energetically-weighted version of this matrix~\cite{camargo1998,schmid2007,camporeale2010}. Since the energy in a given Fourier mode is

@@

\beqar
\label{fourier_en_eqn}
\frac{\partial}{\partial t} \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right) = \mathbf{B}_k \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right) + \mathbf{N}_k \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right), \\ \nonumber \\
\mathbf{B}_k = \left( \begin{array}{cc} -\xi k_z^2 - D k_\perp^4 & -i \kappa \frac{k_y}{k_\perp} + \xi \frac{k_z^2}{k_\perp} \\  \xi \frac{k_z^2}{k_\perp} & - \xi \frac{k_z^2}{k_\perp^2} - D k_\perp^4\end{array} \right)
\eeqar
Since the square $L_2$-norm of the new state vector $\frac{1}{\sqrt{2}} \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right)$ gives the energy contained in the mode $k$,
$\mathbf{B}_k$ is the energetically-weighted matrix of interest. In the limit of large $\xi k_z^2$ as long as $k_\perp$ is also not too large:

@@

\begin{figure}
\centerline{\includegraphics[width=0.45\textwidth]{max_en_growth}}
\caption{$G_{k,\alpha,\rm{max}}(t) = ||e^{\mathbf{B}_{k,\alpha} t}||^2$ for $k_x=0, k_y=1$ for different values of the 2D adiabaticity parameter 
$\alpha$ compared to the energy growth $e^{2 \gamma_{s,\alpha,k} t}$ of the least stable eigenmode of  $\mathbf{B}_{k,\alpha}$. 
The time axis is normalized to $\gamma_{s,\alpha}$, which is different for each of the curves. This normalization makes the $e^{2 \gamma_{s,\alpha,k} t}$ line the same for all $\alpha$. 
$||e^{\mathbf{B}_{k,\alpha} t}||^2$ displays large non-exponential growth at early times that increases with decreasing $\alpha$.}
\label{max_en_growth}
\end{figure}

@@

Linear non-normality is important for the effect it has on turbulence. It is most often cited as the sufficient condition for turbulence in subcritical systems, 
but it can also impact systems that have unstable linear eigenmodes like the HW system. Since certain superpositions of the linear eigenmodes can inject energy faster than the most unstable linear
eigenmode, there is the possibility that the energy injection into the turbulence is faster and at different wavenumbers than that suggested by linear eigenmode analysis. 
To understand this concept and show that it occurs in the nonlinear HW model, we compare the fastest growing linear eigenmode growth rate $\gamma_{s,k}$ to the turbulent growth rate $\gamma_{t,k}$. 
First, however, we define the turbulent growth rate and show how to calcuate it for the HW model.
In order to find the turbulent growth rate for the HW model, we use a Fourier-decomposed energy dynamics analysis~\cite{camargo1995,friedman2012b,friedman2013}. 
To do this, we take Eqs.~\ref{n_eq} and~\ref{phi_eq} and substitute the following:

\beqar
\label{subs_fourier}
n = \sum_{k} n_k {\rm sin} \left( k_x x \right) e^{i k_y y + i k_z z} \\
\phi = \sum_{k} \phi_k {\rm sin} \left( k_x x \right) e^{i k_y y + i k_z z},
\eeqar
where again $k$ is shorthand for $(k_x,k_y,k_z)$.
Then, we multiply the resulting first equation by $n^*_{k'} {\rm sin} \left( k'_x x \right) e^{-i k'_y y - i k'_z z}$ and the second equation by 
$-\phi^*_{k'} {\rm sin} \left( k'_x x \right) e^{-i k'_y y - i k'_z z}$, take the volume integral and add the equations together. The result is

\beqar
\label{Explicit_En_eqn}
\frac{1}{2} \diff{}{t} \left( |n_k|^2 + k^2_\perp |\phi_k|^2 \right) = -i k_y \kappa \phi_k n^*_k - \xi k_z^2 |n_k - \phi_k|^2 \nonumber \\
- D k^4_\perp \left( |n_k|^2 + k^2_\perp |\phi_k|^2 \right) + \sum_{k'} T(k,k')
\eeqar
where, again, the energy $E_k$ is $\frac{1}{2} \left( |n_k|^2 + k^2_\perp |\phi_k|^2 \right)$, 
and the term $\sum_{k'} T(k,k')$ represents the nonlinear contribution (three-wave transfers between different $k$'s), 
which we do not write explicitly. This energy evolution equation may be written symbolically as

\beq
\label{dEdt_def}
\diff{E_k}{t} = \diff{E_{l,k}}{t} + \diff{E_{nl,k}}{t}
\eeq
where $\diff{E_{l,k}}{t}$ represents the terms that come from the linear terms in Eqs.~\ref{n_eq} and~\ref{phi_eq}. These are actually quadratic in the fluctuating
quantities $n_k$ and $\phi_k$ in Eq.~\ref{Explicit_En_eqn}. 
$\diff{E_{l,k}}{t}$ represents the injection of energy into the fluctuations from the free energy in the equilibrium gradients plus the dissipation from collisions and hyperdiffusion.
$\diff{E_{nl,k}}{t}$ represents $\sum_{k'} T(k,k')$ in  Eq.~\ref{Explicit_En_eqn} and comes from the nonlinear terms in 
Eqs.~\ref{n_eq} and~\ref{phi_eq}. The terms in $T(k.k')$ are triadic in the fluctuating quantities and
account for the energy exchange between fluctuations with different $k$. Furthermore, it is conservative: $\sum_{k} \diff{E_{nl,k}}{t} = \sum_{k,k'} T(k,k') = 0$.
Now, in quasi-steady state turbulence, the rate of net energy injection (or dissipation) into the fluctuations at each $k$ by the linear terms must be balanced by
the rate of energy removal (or deposition) from the nonlinear terms. This may be represented formally using growth rates:

@@

We now associate the linear rate of energy injection into the fluctuations $\gamma_{l,k}$ as the turbulent growth rate $\gamma_{t,k}$ discussed above. 
It should be clear now that this growth rate is the turbulence equivalent to the linear eigenmode growth rate $\gamma_{s,k}$~\cite{friedman2012b,terry2006b}. 
Moreover, $\gamma_{l,k}$ is calculated from the spatial structures of the plasma state variables, so it is always well-defined, even in a turbulent plasma. In a linear simulation run for a
sufficiently long time where transients die away, the calculation of $\gamma_{l,k}$ gives exactly $\gamma_{s,k}$.

In a turbulent state, which can be considered a linear superposition of linear eigenvectors, $\gamma_{t,k} \le \gamma_{s,k}$ if the system is normal. Recall that in normal systems, the maximal
energy growth is given by the most unstable linear eigenmode. In other words, maximal growth is achieved when the most stable eigenmode has zero amplitude. When this subdominant eigenmode has finite
amplitude, the growth rate must be lowered. On the other hand, non-normal systems may grow faster than is dictated by the most unstable eigenmode when the subdominant eigenmode has finite amplitude. 
It is therefore possible for $\gamma_{t,k} \ge \gamma_{s,k}$ in non-normal systems.

\begin{figure}
\centerline{\includegraphics[width=0.55\textwidth]{alpha1e-5_gamma_s_t}}
\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$ and b) the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 10^{-5}$. Note the different scales.}
\label{alpha1e-5_gamma_s_t}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.55\textwidth]{alpha1_gamma_s_t}}
\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$ and b) the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 1$.}
\label{alpha1_gamma_s_t}
\end{figure}

For the 2D HW model, we show
$\gamma_{s,k}$ and $\gamma_{t,k}$ for two different values of $\alpha$. In Fig.~\ref{alpha1e-5_gamma_s_t}, we show these growth rate spectra for $\alpha = 10^{-5}$, for which the system is
highly non-normal, and in Fig.~\ref{alpha1_gamma_s_t} for $\alpha = 1$, for which the system is very close to being normal. Note the different scales used in all of the figures. In the highly
non-normal case, the turbulent growth rate is somewhat higher than the eigenmode growth rate for most values of $k_x,k_y$, and about 6 times higher where the spectrum peaks. For the more normal
case, in contrast, $\gamma_{t,k}$ is generally lower than $\gamma_{s,k}$, with the peak at less than a factor of 2 less, though the peaks are at slightly different values of $k_y$.

\begin{figure}
\centerline{\includegraphics[width=0.4\textwidth]{gamma_max_vs_alpha}}
\caption{}
\label{gamma_max_vs_alpha}
\end{figure}

We show the peak values for $\gamma_{s,k}$ and $\gamma_{t,k}$ in Fig.~\ref{gamma_max_vs_alpha} as a function of $\alpha$. The peak values always occur in $k$-space
at the lowest $k_x$ available to our system, corresponding to half of a period of a sine wave. The value of $k_y$ that gives the highest growth rate migrates upward as $\alpha$ increases,
as is clear from Figs.~\ref{alpha1e-5_gamma_s_t} and~\ref{alpha1_gamma_s_t}. Furthermore, as expected, when $\alpha$ is small $\lesssim 10^{-3}$, $\gamma_{t,k} > \gamma_{s,k}$, with
$\gamma_{t,k} \gg \gamma_{s,k}$ as $\alpha \to 0$. Additionally, when $\alpha$ becomes larger and the system becomes more normal ($\gtrsim 10^{-1}$), $\gamma_{t,k} < \gamma_{s,k}$.
For both the normal and highly non-normal limits of the 2D HW model, $\gamma_{s,k}$ does a poor job of predicting $\gamma_{t,k}$ because $\gamma_{s,k}$ neglects the stable branch of the
dispersion relation -- the damped eigenmode. The importance of the stable eigenmode branch has started to gain acceptance in the plasma community, although it has generally been used to explain
turbulent saturation, as is the case for nearly normal systems~\cite{terry2006}. The increased turbulent excitation in the highly non-normal case, however, is not as well-known but it also
important.

\begin{figure}
\centerline{\includegraphics[width=0.4\textwidth]{gamma_max_vs_kz}}
\caption{}
\label{gamma_max_vs_kz}
\end{figure}

The nonlinear simulation reveals a fascinating property of the turbulence -- 
it is dominated by a nonlinear instability process despite being linearly unstable to drift waves~\cite{friedman2012b,friedman2013}.
The nonlinear instability, which was discovered by Drake et al.~\cite{drake1995}, works as follows: 
magnetic-field-aligned ($k_\para=0$) convective filaments transport density across the equilibrium density gradient, setting up $k_\para=0$ density filaments. 
These filaments are unstable to secondary drift waves, which grow on the periphery of the filaments. 
These drift waves, which have finite $k_\para$, nonlinearly couple to one another and generate new convective filaments.

Although the instability is called a nonlinear instability, the first part of the mechanism -- the transport of background density by the convective filaments -- is a linear one.
In fact, the other parts of the mechanism are driven by energetically conservative nonlinear interactions, 
meaning that the convective transport is the only step responsible for energy injection into the fluctuations.

This convective transport of density filaments is akin to the paradigmatic ``lift-up'' mechanism in hydrodynamic shear flows whereby streamwise vortices drive streamwise 
streaks~\cite{trefethen1993,krommes1999}.
Both are transient growth processes.

\section{Non-Modal Linear Technique of Turbulent Growth Rate}

>>>>

Turbulence in magnetically confined plasmas is almost always attributed to linear instabilities. The calculation of linear instabilities through normal mode analysis
-- the calculation of eigenvalues and eigenvectors of a linearized dynamical system -- has successfully predicted the onset criterion for many MHD and micro-instabilities.
Nevertheless, it is well-known in the hydrodynamics community that normal mode analysis fails at predicting turbulent onset for a large class of instabilities that are called subcritical.
Subcritical turbulence is certainly possible in plasmas, and several specific cases have been studied in the past~\cite{camargo1998,camporeale2010,schekochihin2012}. 
For the most part, however, normal mode analysis dominates plasma turbulence prediction.

Since the hydrodynamics community has long studied subcritical turbulence, it has much to offer in terms of understanding the phenomenon. Most importantly,
in the early 1990's, Trefethen and others attributed the failure of normal mode analysis in predicting subcritical turbulent onset to the non-normality of linear operators of
dynamical systems~\cite{trefethen1993,schmid2007}. A non-normal operator has 
eigenvectors that are not orthogonal to one another. One consequence of eigenvector nonorthogonality is that even when all eigenvectors decay exponentially under linear evolution, 
superpositions of eigenvectors can grow, albeit transiently.
In other words, certain fluctuations of the laminar state can access free energy from background gradients even though normal mode fluctuations cannot.
When combined with nonlinear effects, this allows for sustained subcritical turbulence by a bootstrapping process.
Such behavior is obscured by traditional normal mode analysis, which only effectively describes the long time asymptotic behavior of fluctuations under  
action of the linear operator. Transient growth, which can dominate turbulent evolution, can be discovered only through non-modal calculations.

Non-modal analysis generally employs tools such as pseudospectra and maximum possible transient growth curves, which have been used to successfully explain why turbulence is subcritical in various
systems. However, to our knowledge, they have not been used to predict onset criteria in subcritical systems (e.g. the transition Reynold's number in pipe flows).

Furthermore, non-modal treatments have generally been explanatory rather than predictive and have centered around the transition to turbulence in subcritical systems rather 
than on properties of fully-developed turbulence.
This paper takes up the task of developing an approach to understand turbulent properties using only non-modal linear calculations with the goal of 
ultimately making quantitative predictions.  Our approach is to calculate an average growth rate of turbulent fluctuations due to linear processes, specifically due to transient growth.  We model the
turbulent steady state as a series of processes:  (1) the turbulence starts as a spatially random state, (2) linear transient growth deterministically amplifies the turbulent energy (or
decrease it in wavenumber ranges where linear damping dominates), (3) nonlinear transfer sets in at a specified timescale, terminating the
transient growth process and re-randomizing the turbulent state (at which point the cycle repeats).    Optimally, the timescale for the final step would be the
nonlinear decorrelation time of the turbulent system, but in order to enable predictive capability, we employ critical balance
arguments to use a characteristic \emph{linear} time. Since there is no obvious single linear time scale, we test several and compare the results to determine which works best.
The procedure ultimately produces a growth rate that can be used
to predict turbulent properties such as saturation levels and transport rates through mixing length arguments.

@@

\beqar
\label{n_eq}
\pdiff{n}{t} = - {\mathbf v_E} \cdot \grad n - \kappa \pdiff{\phi}{y} + \xi \gradpar^2 (n - \phi) - D \gradperp^4 n, \\
\label{phi_eq}
\pdiff{\gradperp^2 \phi}{t} = - {\mathbf v_E} \cdot \grad (\gradperp^2 \phi) + \xi \gradpar^2 (n - \phi) - D \gradperp^6 \phi
\eeqar
The normalizations are $\phi \to e \phi/T_e, n \to n/n_0, \grad \to \rho_s \grad , x,y,z \to x,y,z/\rho_s, t \to \omega_{ci} t $. Additionally,
$\mathbf{v_E} = \mathbf{b} \times \grad \phi$, $\xi = 4 \pi \omega_{ce}/\nu_{ei}$, 
and $\kappa = \rho_s/L_n$, where $L_n$ is the density scale length. The terms $D \gradperp^4 n$ and $D \gradperp^6 \phi$ are artificial hyper-diffusion and hyper-viscosity used to dissipate
energy on small scales. We set $D \to 10^{-5}$ except when otherwise indicated.
We associate the direction along $\mathbf{B}$ as the parallel or $z$ direction. The direction perpendicular $\mathbf{B}$ that
points opposite the density gradient is defined as the $x$ direction, and the $y$ direction is taken perpendicular to both of these. 
In the analysis and numerical simulations, which we perform with the BOUT++ code~\cite{dudson2009}, we use periodic boundary conditions in
the $y$ and $z$ directions with zero-value boundaries in the $x$ direction. For the 2D version of the model, we replace $-\xi \gradpar^2$ by 
$\alpha = k_z^2 \xi = 4 \pi k_z^2 \rho_s^2 \omega_{ce} /\nu_{ei}$, which is commonly known as the adiabaticity parameter~\cite{camargo1995,camargo1998}.
In nonlinear simulations, the turbulence can laminarize due to the build-up of $k_y=0$ density and potential structures~\cite{biskamp1995}. To prevent this, we subtract out the $k_y=0$
$n$ and $\phi$ components in the simulations, which is the equivalent of adding source terms to Eqs.~\ref{n_eq} and~\ref{phi_eq}~\cite{friedman2012b}.

@@

\beqar
\label{fourier_eqn}
\frac{\partial}{\partial t} \left( \begin{array}{cc} n_k \\ \phi_k \end{array} \right) = \mathbf{A}_k \left( \begin{array}{cc} n_k \\ \phi_k \end{array} \right) + \mathbf{N}_k \left( \begin{array}{cc} n_k \\ \phi_k \end{array} \right), \\ \nonumber \\
\label{A_k}
\mathbf{A}_k = \left( \begin{array}{cc} -\xi k_z^2 - D k_\perp^4 & -i \kappa k_y + \xi k_z^2 \\  \xi \frac{k_z^2}{k_\perp^2} & - \xi \frac{k_z^2}{k_\perp^2} - D k_\perp^4\end{array} \right)
\eeqar
with $\mathbf{A}_k$ being the linear coupling matrix and $\mathbf{N}_k$ the nonlinear matrix, which we don't write explicitly. 
Now in assessing the degree of normality of the HW model, we are not interested in the normality of $\mathbf{A}_k$ itself. 
Rather, we are interested in the energetically-weighted version of this matrix~\cite{camargo1998,schmid2007,camporeale2010}. Since the energy in a given Fourier mode is

@@

\beqar
\label{fourier_en_eqn}
\frac{\partial}{\partial t} \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right) = \mathbf{B}_k \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right) + \mathbf{N}_k \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right), \\ \nonumber \\
\label{B_k}
\mathbf{B}_k = \left( \begin{array}{cc} -\xi k_z^2 - D k_\perp^4 & -i \kappa \frac{k_y}{k_\perp} + \xi \frac{k_z^2}{k_\perp} \\  \xi \frac{k_z^2}{k_\perp} & - \xi \frac{k_z^2}{k_\perp^2} - D k_\perp^4\end{array} \right)
\eeqar
Since the square $L_2$-norm of the new state vector $\frac{1}{\sqrt{2}} \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right)$ gives the energy contained in the mode $k$,
$\mathbf{B}_k$ is the energetically-weighted matrix of interest. In the limit of large $\xi k_z^2$ as long as $k_\perp$ is also not too large:

@@

\begin{figure}
\centerline{\includegraphics[width=0.45\textwidth]{max_en_growth}}
\caption{$G_{k,\alpha,\rm{max}}(t) = ||e^{\mathbf{B}_{k,\alpha} t}||^2$ for $k_x=0, k_y=1, \kappa=1$ for different values of the 2D adiabaticity parameter 
$\alpha$ compared to the energy growth $e^{2 \gamma_{s,\alpha,k} t}$ of the least stable eigenmode of  $\mathbf{B}_{k,\alpha}$. 
The time axis is normalized to $\gamma_{s,\alpha}$, which is different for each of the curves. This normalization makes the $e^{2 \gamma_{s,\alpha,k} t}$ line the same for all $\alpha$. 
$||e^{\mathbf{B}_{k,\alpha} t}||^2$ displays large non-exponential growth at early times that increases with decreasing $\alpha$.}
\label{max_en_growth}
\end{figure}

@@

Linear non-normality is important for the effect it has on turbulence. It is most often cited as the sufficient condition for turbulence in subcritical systems, 
but it can also impact systems that have unstable linear eigenmodes like the HW system. Since certain superpositions of the linear eigenmodes can inject energy faster than the most unstable linear
eigenmode, there is the possibility that the energy injection into the turbulence is faster and at different wavenumbers than is suggested by linear eigenmode analysis. 
To understand this concept and show that it occurs in the nonlinear HW model, we compare the fastest growing linear eigenmode growth rate $\gamma_{s,k}$ to the turbulent growth rate $\gamma_{t,k}$. 
To define the turbulent growth rate in general and specifically for the HW model, we employ a Fourier-decomposed energy dynamics analysis~\cite{camargo1995,friedman2012b,friedman2013}. 
To do this, we take Eqs.~\ref{n_eq} and~\ref{phi_eq} and substitute the following:

\beqar
\label{subs_fourier}
n = \sum_{k} n_k \ {\rm sin} \left( k_x x \right) e^{i k_y y + i k_z z} \\
\phi = \sum_{k} \phi_k \ {\rm sin} \left( k_x x \right) e^{i k_y y + i k_z z},
\eeqar
where, again, $k$ stands for $(k_x,k_y,k_z)$.
Then, we multiply the resulting first equation by $n^*_{k'} \ {\rm sin} \left( k'_x x \right) e^{-i k'_y y - i k'_z z}$ and the second equation by 
$-\phi^*_{k'} \ {\rm sin} \left( k'_x x \right) e^{-i k'_y y - i k'_z z}$, take the volume integral and add the two equations together. The result is

\beqar
\label{Explicit_En_eqn}
\frac{1}{2} \diff{}{t} \left( |n_k|^2 + k^2_\perp |\phi_k|^2 \right) = -i k_y \kappa \phi_k n^*_k - \xi k_z^2 |n_k - \phi_k|^2 \nonumber \\
- D k^4_\perp \left( |n_k|^2 + k^2_\perp |\phi_k|^2 \right) + \sum_{k'} T(k,k'), \quad \quad
\eeqar
where, again, the energy $E_k$ is $\frac{1}{2} \left( |n_k|^2 + k^2_\perp |\phi_k|^2 \right)$, 
and the term $\sum_{k'} T(k,k')$ represents the nonlinear contribution (three-wave transfers between different $k$'s), 
which we do not write explicitly. This energy evolution equation may be written symbolically as

\beq
\label{dEdt_def}
\diff{E_k}{t} = \diff{E_{l,k}}{t} + \diff{E_{nl,k}}{t}
\eeq
where $\diff{E_{l,k}}{t}$ represents the terms that come from the linear terms in Eqs.~\ref{n_eq} and~\ref{phi_eq}, which are quadratic in the fluctuating
quantities $n_k$ and $\phi_k$ in Eq.~\ref{Explicit_En_eqn}. 
$\diff{E_{l,k}}{t}$ represents the injection of energy into the fluctuations from the free energy in the equilibrium gradients plus the dissipation from collisions and hyperdiffusion.
$\diff{E_{nl,k}}{t}$ represents $\sum_{k'} T(k,k')$ in  Eq.~\ref{Explicit_En_eqn} and comes from the nonlinear terms in 
Eqs.~\ref{n_eq} and~\ref{phi_eq}. The terms in $T(k,k')$ are triadic in the fluctuating quantities and
account for the energy exchange between fluctuations with different $k$'s. Furthermore, it is conservative: $\sum_{k} \diff{E_{nl,k}}{t} = \sum_{k,k'} T(k,k') = 0$.

Now, in quasi-steady state turbulence, the rate of net energy injection (or dissipation) into the fluctuations at each $k$ by the linear terms must be balanced by
the rate of energy removal (or deposition) from the nonlinear terms. This may be represented formally using growth rates:

@@

We now associate the turbulent growth rate $\gamma_{t,k}$ discussed above with the linear rate of energy injection into the fluctuations $\gamma_{l,k}$ because $\gamma_{l.k}$
determines how fast and where in wavenumber space energy is injected into the turbulence.
$\gamma_{l,k}$, in other words, is the turbulence equivalent to the linear eigenmode growth rate $\gamma_{s,k}$~\cite{friedman2012b,terry2006b}. 
That is, in a linear simulation run for a sufficiently long time where transients die away, the calculation of $\gamma_{l,k}$ gives exactly $\gamma_{s,k}$.
Moreover, $\gamma_{l,k}$ is calculated from the spatial structures of the plasma state variables, so it is always well-defined, even in a turbulent plasma.

A turbulent system, which can be decomposed into a linear eigenvector basis, has $\gamma_{t,k} \le \gamma_{s,k}$ if the system is normal. Recall that in normal systems, the maximal
energy growth is given by the most unstable linear eigenmode. In other words, maximal growth is achieved when the \emph{most stable} eigenmode has zero amplitude.
On the other hand, non-normal systems may grow faster when the more stable eigenmode has finite amplitude than when it has none (see Fig.~\ref{max_en_growth}). 
It is therefore possible for $\gamma_{t,k} > \gamma_{s,k}$ in non-normal systems.

\begin{figure}
\centerline{\includegraphics[width=0.52\textwidth]{alpha1e-5_gamma_s_t}}
\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$ and b) the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 10^{-5}, \kappa=1$. Note the different scales.}
\label{alpha1e-5_gamma_s_t}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.52\textwidth]{alpha1_gamma_s_t}}
\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$ and b) the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 1, \kappa=1$.}
\label{alpha1_gamma_s_t}
\end{figure}

For the 2D HW model, we show
$\gamma_{s,k}$ and $\gamma_{t,k}$ for two different values of $\alpha$. In Fig.~\ref{alpha1e-5_gamma_s_t}, we show these growth rate spectra for $\alpha = 10^{-5}$, for which the system is
highly non-normal, and in Fig.~\ref{alpha1_gamma_s_t} for $\alpha = 1$, for which the system is very close to being normal. All calculations in this section use $\kappa=1$. 
Note the different scales used in all of the figures. In the highly
non-normal case, the turbulent growth rate is somewhat higher than the eigenmode growth rate for most values of $k_x,k_y$, and about 6 times higher where the spectrum peaks. For the more normal
case, in contrast, $\gamma_{t,k}$ is generally lower than $\gamma_{s,k}$, with the peak at nearly a factor of 2 less, with the peaks being at slightly different values of $k_y$.

\begin{figure}
\centerline{\includegraphics[width=0.4\textwidth]{gamma_max_vs_alpha}}
\caption{The eigenmode $\gamma_{s}$ and turbulent $\gamma_{t}$ growth rates as a function of $\alpha$, which controls the normality of the linear operator. All growth rates are peaks in $k_x-k_y$ space.}
\label{gamma_max_vs_alpha}
\end{figure}

We show the peak values for $\gamma_{s,k}$ and $\gamma_{t,k}$ in Fig.~\ref{gamma_max_vs_alpha} as a function of $\alpha$. The peak values always occur in $k$-space
at the lowest $k_x$ available to our system, corresponding to half of a period of a sine wave. The value of $k_y$ that gives the highest growth rate migrates upward as $\alpha$ increases,
as is clear from Figs.~\ref{alpha1e-5_gamma_s_t} and~\ref{alpha1_gamma_s_t}. Furthermore, as expected, when $\alpha$ is small ($\lesssim 10^{-3}$), $\gamma_{t,k} > \gamma_{s,k}$, with
$\gamma_{t,k} \gg \gamma_{s,k}$ as $\alpha \to 0$. Additionally, when $\alpha$ becomes larger and the system becomes more normal ($\gtrsim 10^{-1}$), $\gamma_{t,k} < \gamma_{s,k}$.

For both the normal and highly non-normal limits of the 2D HW model, $\gamma_{s,k}$ does a poor job of predicting $\gamma_{t,k}$ because $\gamma_{s,k}$ neglects the stable branch of the
dispersion relation -- the damped eigenmode. The importance of the stable eigenmode branch (or branches) 
has started to gain acceptance in the plasma community, although it has generally been used in normal
or nearly normal systems to explain turbulent saturation~\cite{terry2006b} ($\gamma_{t,k} < \gamma_{s,k}$). The case of highly non-normal turbulence and the associated increased turbulent excitation
hasn't been understood as well. In fact, simulation and analysis of the 3D HW model (and extended models) is an area in which others have seen the effects of strong non-normal transient growth
but have not realized them as such.

\begin{figure}
\centerline{\includegraphics[width=0.4\textwidth]{gamma_max_vs_kz}}
\caption{The growth rates of the fastest growing eigenmode $\gamma_{s,k}$ and the of the turbulence $\gamma_{t,k}$ as a function of $k_z$ for the 3D HW model. The most remarkable feature is
that the turbulent growth rate peaks at $k_z=0$ despite the stability of all linear eigenmodes at $k_z=0$.}
\label{gamma_max_vs_kz}
\end{figure}

The 3D HW model nonlinear simulation has a fascinating property -- the turbulent energy injection is strongest at $k_z = 0$ despite the fact that there can be no unstable linear eigenmodes at $k_z=0$.
We show this in Fig.~\ref{gamma_max_vs_kz} using the eigenmode $\gamma_{s,k}$ and turbulent $\gamma_{t,k}$ growth rates, plotting them as a function of $k_z$.
The long-stated cause of this $k_z=0$ energy injection is a nonlinear instability, which was discovered by Biaskamp and Zeiler~\cite{biskamp1995} and explored further by Drake et al.~\cite{drake1995}. 
The nonlinear instability works as follows: magnetic-field-aligned ($k_z=0$) convective cells transport density across the equilibrium density gradient, setting up $k_z=0$ density structures. 
These structures are unstable to drift waves, which grow as a secondary instability.
The secondary drift waves, which have finite $k_z$, nonlinearly couple to one another and reinforce the original convective cells (or generate new ones), leading to self-sustainment.
